---
title: "Chapter 01: Basic Data Manipulation. Subsetting and filtering."
author: "Aleksandr Taraskin"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Chapter 01: Basic Data Manipulation. Subsetting and filtering.

## 1.1 Subsetting Data

Subsetting data allows you to extract specific rows and columns from
your dataset. This is useful when you want to focus only on the data
that interests you.

Available datasets in R by default

```{r}
# List all available datasets
data()
```

One of the standard learning data sets included in R is the “ToothGrowth” data set. The tooth growth data set is the length of the odontoblasts (teeth) in each of 10 guinea pigs at three Vitamin C dosage levels (0.5, 1, and 2 mg) with two delivery methods (orange juice or ascorbic acid).

The file contains 60 observations of 3 variables

    len : Tooth length
    supp : Supplement type (VC or OJ)
    dose : Dose in milligrams


```{r}
# Load the built-in mtcars dataset
data(ToothGrowth)
# Rename it to look more concise
dataset <- ToothGrowth
# Display the first 6 (default) rows
head(dataset)
```

### 1.1.1 Extracting Columns

List columns
```{r}
colnames(dataset)
```
List of rows
```{r}
rownames(dataset)
```

List columns
```{r}
colnames(dataset)
```

To extract one or more columns from a dataframe, you can use square
brackets `[ ]`.

```{r}
dataset["supp"]
```


What are the options for "supp" column values?

```{r}
unique(dataset["supp"])
```

More convenient way to do that

```{r}
table(dataset["supp"])
```

We can get values only from specific column (different output).
In order to do that we nee to use double brackets: `[[ ]]`

```{r}
dataset[["supp"]]
```

Another syntax for that:
```{r}
dataset$supp
```
`dataset[["supp"]]` and `dataset$supp` are identical
`dataset["supp"]` gives us a completely different output!!!

Different from
```{r}
dataset["supp"]
```

What is "Levels"?

```{r}
levels(dataset$supp)
```
No levels in "dose" column. Why?
```{r}
levels(dataset$dose)
```
Three unique value for "dose" column
```{r}
table(dataset["dose"])
```
They are just numbers, not a "factor" type
```{r}
dataset$dose
```

Let's create another dataset and transform its "dose" column into "factor" type
```{r}
dataset_altered <- dataset
dataset_altered$dose <- factor(dataset_altered$dose)
table(dataset_altered["dose"])
```
Now we have levels!
```{r}
levels(dataset_altered$dose)
```

```{r}
dataset_altered$dose
```



Alternatively, to subset any column you can use the index:

```{r}
dataset[, 2]
```

Wait. What will happen if we put the first index instead?
We will subset a row!
```{r}
dataset[2 , ]
```
If we specify both the first and the second index - we get the value!

```{r}
dataset[1,2]
```


```{r}
dataset[2,3]
```

Double brackets won't change anything there. We get the value already.

```{r}
dataset[[2,3]]
```


### 1.1.2 Extracting Rows

You can extract rows based on their indices or logical conditions.

#### Extracting by Indices

# Extract the first 5 rows

```{r}
dataset[1:5, ]
```
No double brackets this time!
```{r}
result <- tryCatch({
  
  # Won't work
  dataset[[1:5, ]] 
  
  
  
}, error = function(e) {
  message("An error occurred: ", e$message)  
  NULL  
})


print(result)
```

Second index can be specified
```{r}
dataset[1:5, 3]
```


#### Logical Subsetting. Filtering.

You can extract rows based on conditions. For example, to extract all
guinea pigs which drank orange juice =)

```{r}
# Condition in this case is in the specific column, but we return corresponding rows
dataset[dataset$supp == "VC", ]
```

At first glance it can be very confusing. Let's break it into steps:

```{r}
dataset$supp
```

```{r}
dataset$supp == "VC"
```

Another way to subset - pass boolean vector!

```{r}
small_example <- c(dataset$supp[1:2], dataset$supp[(length(dataset$supp)-1):length(dataset$supp)])
```


```{r}
print("before")
small_example
```


```{r}
print("after")
small_example[c(TRUE, TRUE, FALSE, FALSE)]
```

Condition for column values gives us boolean vector output for each row that we need to pass as a first "row" index in brackets
```{r}
dataset[dataset$supp == "VC", ]
```

The same for any other condition. We just need the boolean vector:

```{r}
summary(dataset$len)
```
Higher than median
```{r}
dataset$len > 19.25
```
```{r}
dataset[dataset$len > 19.25, ]
```


### 1.1.3 Combining Row and Column Extraction

You can extract both rows and columns simultaneously. 
```{r}
dataset[dataset$len > 19.25, ]$supp
```
OR
```{r}
dataset[dataset$len > 19.25, "supp"]
```
```{r}
dataset[dataset$len > 19.25, c("len", "dose")]
```

Multiple conditions. Logical "AND"
```{r}
dataset[(dataset$len > 19.25 & dataset$dose == 2), c("len", "dose")]
```
Logical "OR"

```{r}
extreme <- dataset[(dataset$len > 25.27 | dataset$len < 13.07), c("len", "dose")]
extreme
```

```{r}
table(extreme$dose)
```

```{r}
extreme[extreme$dose == 2, "len"]
```

```{r}
extreme[extreme$dose == 0.5, "len"]
```

################################################################################
# TASK
################################################################################

What are extreme "len" values for dosage level 1 mg?
```{r}
# extreme <- dataset[(dataset$len > 25.27 | dataset$len < 13.07), c("len", "dose")]
```
```{r}

```


### Conclusion

In this section, we learned how to extract subsets of data from a
dataframe in R. This included extracting individual columns, rows by
indices, and logical conditions, as well as combinations of these
methods.

