---
title: "Chapter 04: Data Wrangling with dplyr: Filtering, grouping, summarizing, and joining datasets."
author: "Aleksandr Taraskin"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Chapter 04: Data Wrangling with dplyr: Filtering, grouping, summarizing, and joining datasets.

## 4.1 Package "dplyr"

The dplyr package in R is a powerful and widely used package designed for data manipulation and transformation. Developed as part of the tidyverse ecosystem, dplyr provides an intuitive and consistent set of functions to work with data frames and other tabular data structures, making it easier to clean, filter, summarize, and transform data. It’s especially popular in data science and analytics because it simplifies common data manipulation tasks.

To use dplyr we need to load it first
```{r}
library(dplyr)
```

Now let's upload our favourite dataset
```{r}
# Load the built-in dataset
data(ToothGrowth)
# Rename it to look more concise
dataset <- ToothGrowth
# Display the first 6 (default) rows
head(dataset)
```


### 4.1.1 Extracting Columns

List columns
```{r}
colnames(dataset)
```
List of rows
```{r}
rownames(dataset)
```

The main feature of dplyr is the "pipe" operator, it looks like that:

`%>%`

Now instead of `dataset["supp"]` let's use "dplyr" package for that
```{r}
output <- dataset %>% 
  select("supp")
output
```

We can operate with output in the same way

```{r}
table(output)
```

But what is the `dataset[["supp"]]` alternative in this case?

```{r}
#  old way
dataset[["supp"]]

# new way
output <- dataset %>% 
  pull("supp")
output
```

In base R we know how to extract column by its index: `dataset[, 2]`
What is the equvalent in "dplyr"
```{r}
# old way
dataset[, 2]

#  new way
output <- dataset %>% 
  pull(2)
output
```

### 4.1.2 Extracting Rows

Now row
```{r}
#  old way
dataset[2, ]

#  new way
output <- dataset %>% 
  slice(2)
output
```

This internal structure `1:5` also called a "slice"

```{r}
#  old way
dataset[1:5, ]

#  new way
output <- dataset %>% 
  slice(1:5)
output
```


Now both indices are used

```{r}
#  old way
dataset[1,2]


#  new way
output <- dataset %>% 
  slice(1) %>% 
  pull(2)
output
```

```{r}
#  old way
dataset[2,3]


#  new way
output <- dataset %>% 
  slice(2) %>% 
  pull(3)
output
```


```{r}
#  old way
dataset[1:5, 3]


#  new way
output <- dataset %>% 
  slice(1:5) %>% 
  pull(3)
output
```

#### 4.2 Logical Subsetting. Filtering.

Extracting all guinea pigs which drank orange juice

```{r}
# Condition in this case is in the specific column, but we return corresponding rows
# old way
dataset[dataset$supp == "VC", ]

# new way
output <- dataset %>% 
  filter(supp == "VC")
output
```

The same for any other type of conditions.

```{r}
summary(dataset$len)
```
Higher than median
```{r}
# old way
dataset[dataset$len > 19.25, ]
# new way
output <- dataset %>% 
  filter(len > 19.25)
output
```

Now where the "pipe" gets useful:

```{r}
# old way
dataset[dataset$len > 19.25, c("len", "dose")]

# new way

# new way
output <- dataset %>% 
  filter(len > 19.25) %>% 
  select(c("len", "dose"))
output
```

Multiple conditions. Logical "AND"
```{r}
#  old way
dataset[(dataset$len > 19.25 & dataset$dose == 2), c("len", "dose")]

#  new way
output <- dataset %>% 
  filter(len > 19.25 & dose == 2) %>% 
  select(c("len", "dose"))
output
```

Logical "OR"

```{r}
# old way
extreme <- dataset[(dataset$len > 25.27 | dataset$len < 13.07), c("len", "dose")]
extreme

# new way
output <- dataset %>% 
  filter(dataset$len > 25.27 | dataset$len < 13.07) %>% 
  select(c("len", "dose"))
output
```

################################################################################
# TASK
################################################################################

What are extreme "len" values for dosage level 1 mg?
```{r}
#  old way
extreme <- dataset[(dataset$len > 25.27 | dataset$len < 13.07) & dataset$dose == 2, c("len", "dose")]
extreme
```

Write your answer here:
```{r}
# new way
output <- dataset %>%
  filter() %>%
  select()
output
```

### 4.3 Sort and order functions alternatives in "dplyr"

In dplyr, the functions arrange() and desc() are used as alternatives to sort() in base R, function order() is not needed but can be used independently as it is. arrange() and desc() functions allow you to easily sort rows of a data frame based on one or more columns.

The arrange() function in dplyr sorts the rows of a data frame based on specified columns in ascending order by default.

```{r}
#  old way
sort(dataset$len)

#  new way 
output <- dataset %>% 
  arrange(len) %>% 
  pull(len)
output
```

To sort in descending order, use desc() within arrange()

```{r}
#  old way
sort(dataset$len, decreasing = TRUE)

#  new way 
output <- dataset %>% 
  arrange(desc(len)) %>% 
  pull(len)
output
```

Sorting by Multiple Columns

You can sort by multiple columns by listing them in arrange(). Sorting will proceed in the order of the columns specified.

```{r}
# ola way
dataset[order(dataset$dose, dataset$supp, -dataset$len), ]

# new way
output <- dataset %>% 
  arrange(dose, supp, desc(len))
output
```

################################################################################
# TASK
################################################################################
Combine filter() and arrange() functions to filter "len" column values lower than 10 and sort them in descending order

```{r}

```


### Conclusion

In this document, we compared two approaches of data manipulation in R: using base R and "dplyr" package. We compared how to extract subsets of data from a dataframe in R. This included extracting individual columns, rows by indices, and logical conditions, as well as combinations of these methods. We learned how to sort vector/(dataset based on any column) in ascending and descending order. Then demonstrated how to sort by multiple columns, even if more than two with both ways!
